---
title: "01_metaplots.Rmd"
author: "Lucy VK"
date: "4/26/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
library(ggplot2)
source("../../util/my_class_functions.R")
source("../../util/_setup.R")

load("../01_consensus_peaks/results/peak_features.RData", verbose = T)

```

# establishing lncRNA and mRNA promoters (+/- 1kb)
```{r create lncRNA and mRNA promoters }

# creating promoter lists
lncrna_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_type == "lncRNA"]
mrna_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_type == "protein_coding"]
super_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_id %in% peak_occurence_df[peak_occurence_df$super_binder == "super binder",]$gene_id]
non_super_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_id %in% peak_occurence_df[peak_occurence_df$super_binder == "not super binder",]$gene_id]

```

# metaplots for each DBP by lncRNA and mRNA promoters

A handful of metaplots look different between lncRNA and mRNA promoters including H3K36me3, ZNF460, H3K4me2 but most look the same.
```{r}

#setting up lncrna df
lncrna_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())

# for loop to populate DF with overlap density in lncrna promoters
for(i in 1:length(filtered_consensus_list)) {
  print(names(filtered_consensus_list)[[i]])
  tmp_df <- profile_tss(filtered_consensus_list[[i]], lncrna_mrna_promoters = lncrna_promoters)
  tmp_df$dbp <- names(filtered_consensus_list)[[i]]
  lncrna_metaplot_df <- bind_rows(lncrna_metaplot_df, tmp_df)
  
}

# saving
write_rds(lncrna_metaplot_df, "results/lncRNA_metaplot_df_final.rds")

# now for mRNAs 
mrna_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())

# for loop to populate mRNA_metaplot
for(i in 1:length(filtered_consensus_list)) {
  print(names(filtered_consensus_list)[[i]])
  tmp_df <- profile_tss(filtered_consensus_list[[i]], lncrna_mrna_promoters = mrna_promoters)
  tmp_df$dbp <- names(filtered_consensus_list)[[i]]
  mrna_metaplot_df <- bind_rows(mrna_metaplot_df, tmp_df)
  
}

# saving mRNA metaplots
write_rds(mrna_metaplot_df, "results/mrna_metaplot_df_final.rds")

# now adding the information of gene type
mrna_metaplot_df$gene_type <- "mRNA"
lncrna_metaplot_df$gene_type <- "lncRNA"
combined_metaplot_profile <- bind_rows(mrna_metaplot_df, lncrna_metaplot_df)

#TODO why not printing
# TODO asign cluster value and plot the average of the clusters
# plotting - NOTE facet wrap by dbp !

# THIS PLOT IS BIGGGG

# pdf(file = "figures/mega_plot_test.pdf")
ggplot(combined_metaplot_profile, 
       aes(x = x, y = dens, color = gene_type )) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  facet_wrap(dbp ~ ., scales = "free_y") +
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-1000, 0, 1000),
                     labels = c("-1kb", "TSS", "+1kb"),
                     name = "") + 
  ylab("Peak frequency") +
 scale_color_manual(values = c("#424242","#a8404c"))

# saving
ggsave("figures/mega_meta_plot_lncRNA-mRNA.pdf", width = 49, height = 12)
# ggsave("figures/mega_meta_plot_lncRNA-mRNA.pdf", limitsize = F)


```
```

Metaplots for super promoters vs. not 
A handful of metaplots look different between super promoters vs. not including E2F2, H3K4me1, ZNF460, PREB, BRCA1, ZKSCAN1 but most look the same.

```{r super binder vs. not metaplot}

#setting up lncrna df
super_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())

# for loop to populate DF with overlap density in lncrna promoters
for(i in 1:length(filtered_consensus_list)) {
  print(names(filtered_consensus_list)[[i]])
  tmp_df <- profile_tss(filtered_consensus_list[[i]], lncrna_mrna_promoters = super_promoters)
  tmp_df$dbp <- names(filtered_consensus_list)[[i]]
  super_metaplot_df <- bind_rows(super_metaplot_df, tmp_df)
  
}

# saving
write_rds(super_metaplot_df, "results/super_metaplot_df_final.rds")

# now for mRNAs 
non_super_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())

# for loop to populate mRNA_metaplot
for(i in 1:length(filtered_consensus_list)) {
  print(names(filtered_consensus_list)[[i]])
  tmp_df <- profile_tss(filtered_consensus_list[[i]], lncrna_mrna_promoters = non_super_promoters)
  tmp_df$dbp <- names(filtered_consensus_list)[[i]]
  non_super_metaplot_df <- bind_rows(non_super_metaplot_df, tmp_df)
  
}

# saving mRNA metaplots
write_rds(non_super_metaplot_df, "results/non_super_metaplot_df_final.rds")

# now adding the information of gene type
super_metaplot_df$gene_type <- "super"
non_super_metaplot_df$gene_type <- "not super"
combined_metaplot_profile <- bind_rows(super_metaplot_df, non_super_metaplot_df)

ggplot(combined_metaplot_profile, 
       aes(x = x, y = dens, color = gene_type )) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  facet_wrap(dbp ~ ., scales = "free_y") +
  ggtitle("Super Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-1000, 0, 1000),
                     labels = c("-1kb", "TSS", "+1kb"),
                     name = "") + 
  ylab("Peak frequency") +
 scale_color_manual(values = c("#424242","#a8404c"))

# SAVE THIS OBJECT


# saving
ggsave("figures/super_mega_meta_plot.pdf", width = 49, height = 12)

```