---
title: "01_clustering"
author: "lucyvk"
date: "4/24/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
library(ggplot2)
library(pheatmap)
source("../../util/my_class_functions.R")
source("../../util/_setup.R")
library(ggdendro)

```

# loading in peak_features env objects from 01_create_consensus_peaks
```{r laod env objects}

load("../01_consensus_peaks/results/peak_features.RData", verbose = T)

```

Create a cluster dendogram based on similarities of DBP binding profiles 
```{r cluster dendogram }

# creating distance matrix
#promoter_peak_occurence_matrix <- as.matrix(promoter_peak_occurrence_matrix)
#peak_occurence_dist <- dist(promoter_peak_occurrence_matrix, method = "binary")
#write_rds(peak_occurence_dist, "results/peak_occurence_dist.rds")
peak_occurence_dist <- readRDS("results/peak_occurence_dist.rds")
# clustering distance values
bin_hier <- hclust(peak_occurence_dist, method = "complete")

pdf("figures/dbp_hclust_dendro.pdf", height = 12, width = 70)
plot(bin_hier)
dev.off()

```
```
We see that there are 5 or six big clusters. In particular, we can look at the locations of some of the ZNF proteins with unknown function. For example, we see ZNF644 clustered in it's own little group with FOXP1 -- maybe they have similar function, so maybe ZNF644 is also involved in transcription regulation. 

We can also look at a GGDENRO version of this clustering:

```{r plot with ggdendro }
 ggdendro::ggdendrogram(bin_hier, rotate = FALSE,  size = 3, 
                       theme_dendro = TRUE) +
   coord_flip() +
   scale_y_continuous() +
   scale_x_continuous(position = "top") +
   scale_x_continuous(breaks = seq_along(bin_hier$labels[bin_hier$order]),
                      
                      # adding labels that are in the order 'column'
             labels = bin_hier$labels[bin_hier$order], position = "top",
             expand = c(0,0)) +
   theme(axis.text.x = element_text(angle = 90, hjust  = 1)) + 
   theme(axis.text.y = element_text(angle = 0,hjust = 1)) +
   scale_y_reverse(expand = c(0.01, 0)) +
   theme(
     plot.background = element_blank(),
     panel.grid.major = element_blank(),
   panel.grid.minor = element_blank(),
     panel.border = element_blank()
   )

ggsave("figures/ggdendro_plot.pdf", height = 50, width = 12, limitsize = F)

```

Now, let's compare the clustering on lncRNA and mRNA promoters separately
```{r lncrna promoter clustering}

lncrna_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_type == "lncRNA"]

lncrna_peak_occurence <- promoter_peak_occurrence_matrix[,lncrna_promoters$gene_id]
bin_hier_lncrna <- hclust(dist(lncrna_peak_occurence, method = "binary"))

ggdendro::ggdendrogram(bin_hier_lncrna, rotate = FALSE,  size = 3, 
                    theme_dendro = TRUE) +
   coord_flip() +
   scale_y_continuous() +
   scale_x_continuous(position = "top") +
   scale_x_continuous(breaks = seq_along(bin_hier_lncrna$labels[bin_hier_lncrna$order]),
                      
                      # adding labels that are in the order 'column'
             labels = bin_hier_lncrna$labels[bin_hier_lncrna$order], position = "top",
             expand = c(0,0)) +
   theme(axis.text.x = element_text(angle = 90, hjust  = 1)) + 
   theme(axis.text.y = element_text(angle = 0,hjust = 1)) +
   scale_y_reverse(expand = c(0.01, 0)) +
   theme(
     plot.background = element_blank(),
     panel.grid.major = element_blank(),
   panel.grid.minor = element_blank(),
     panel.border = element_blank()
)
ggsave("figures/lncrna_hclust_binary_dist.pdf",height = 50, width = 12, limitsize = F)

```

Result -- There appears to be some similarity (e.g. looking at the placement of POLR2A / POLR2AphosphoS2 and FOXA2 / FOXA1), however generally it looks like more of the proteins are in less distinct groups, like maybe the lncRNA promoter binding profiles are more similar across the board.
```{r mrna promoter clustering}
mrna_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_type == "protein_coding"]

mrna_peak_occurence <- promoter_peak_occurrence_matrix[,mrna_promoters$gene_id]
bin_hier_mrna <- hclust(dist(mrna_peak_occurence, method = "binary"))

ggdendro::ggdendrogram(bin_hier_mrna, rotate = FALSE,  size = 3, 
                    theme_dendro = TRUE) +
   coord_flip() +
   scale_y_continuous() +
   scale_x_continuous(position = "top") +
   scale_x_continuous(breaks = seq_along(bin_hier_mrna$labels[bin_hier_mrna$order]),
                      
                      # adding labels that are in the order 'column'
             labels = bin_hier_mrna$labels[bin_hier_mrna$order], position = "top",
             expand = c(0,0)) +
   theme(axis.text.x = element_text(angle = 90, hjust  = 1)) + 
   theme(axis.text.y = element_text(angle = 0,hjust = 1)) +
   scale_y_reverse(expand = c(0.01, 0)) +
   theme(
     plot.background = element_blank(),
     panel.grid.major = element_blank(),
   panel.grid.minor = element_blank(),
     panel.border = element_blank()
)
ggsave("figures/mrna_hclust_binary_dist.pdf", height = 50, width = 12, limitsize = F)

```
Result -- There are again some similarities, although it looks like maybe some of the clusters that were more distinct in the overall clustering are combined.
```